A Type System for String Sanitation\\Implemented Inside a Python

### 2014-04-15 sanitation-plas14/sanitation-plas14.tex
Beginnings of Sec 3 of PLAS paper
Removed <> from syntax
confusion

%TODO-nrf:
% * Type system definition -- check big step semantics
% PRoofs tonigt:
% * Type safety proof for internal language
% * Representational consistency

%\documentclass[12pt]{article}
\documentclass[10pt,preprint]{sigplanconf}
\newcommand{\lamAce}{\lambda_{\text{Ace}}}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{verbatimbox}
\input{../att-icfp14/macros-atlam}
\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{light-gray}{gray}{0.5}
\usepackage{listings}
\usepackage{wasysym}
    \makeatletter
%
% \btIfInRange{number}{range list}{TRUE}{FALSE}
%
% Test if int number <number> is element of a (comma separated) list of ranges
% (such as: {1,3-5,7,10-12,14}) and processes <TRUE> or <FALSE> respectively
%
        \newcount\bt@rangea
        \newcount\bt@rangeb

@@

%
% Begin Grammar 
%
\newcommand{\ereplace}[3]{{\tt replace}[#1](#2,#3)}
\newcommand{\echeck}[2]{{\tt check}[#1](#2)}
\newcommand{\estr}[2]{{\tt str[#1, #2]}}
\newcommand{\estrt}[1]{{\tt strt[#1]}}
\newcommand{\econcat}{+}

\newcommand{\ireplace}[3]{{\tt ireplace}(#1,#2,#3)}
\newcommand{\icheck}[2]{{\tt icheck}(#1, #2)}
\newcommand{\istr}[1]{{{\tt istr}[#1]}}
\newcommand{\istrt}{{\tt istrt}}
\newcommand{\iconcat}{\hat{+}}

@@

@@

<$\tau$> ::=				\hfill	External types					\\
$\tau \rightarrow \tau$ \alt
$\estrt{r}$				\hfill Regular expression types

<e> ::= \hfill External expressions \\
      $x$ \alt
      $\lambda x:\tau.e$ \alt
      $e e$ \alt
      $\estr{r}{s}$ \alt
      $e \econcat e$ \hfill String concatenation \\
      $\ereplace{r}{e}{e}$ \alt
      $\echeck{r}{e}$

<$\sigma$> ::= \hfill Internal types \\
$\tau \rightarrow \tau$ \alt $\istrt$

<i> ::= \hfill Internal expressions \\
  $x$ \alt
  $\lambda x:\sigma.i$ \alt
  $i i$ \alt
  $\istr{s}$ \alt
  $i \econcat i$ \alt
  $\ireplace{r}{i}{i}$ \alt
  $\icheck{r}{i}$
\end{grammar}
\caption{Syntax of $\lcs$}
\label{fig:lcsSyntax}
\end{figure}

@@

\onecolumn
\begin{figure}
\begin{center}
\begin{tabular}{c r}

\inferline{T-Var}
{\gvd x : \tau}
{x : \tau \in \Gamma}

\inferline{T-Abs}
{\gvd \lambda x : \tau .e : \tau \rightarrow \tau' \sreduces \lambda x : \sigma.i}
{\Gamma, x : \tau \vdash e : \tau' \sreduces i \\ \sigma \equiv \trep(\tau)}

\inferline{T-App}
{\Gamma \vdash e_1 e_2 \sreduces i_1 i_2}
{\Gamma \vdash e_1 : \tau \rightarrow \tau' \sreduces i_1 \\ \Gamma \vdash e_2 : \tau \sreduces i_2}

\inferline{T-Replace}
{\gvd \ereplace{r}{e_1}{e_2} : \estrt{r'} \sreduces \ireplace{r}{i_1}{i_2}}
{\gvd e_1 : \estrt{r_1} \ireduces i_1 \\ \gvd e_2 : \estrt{r_2} \ireduces i_2 \\ r' \equiv \crl{r}{r_2}{r_1}}

\inferline{T-Check-Ok}
{\gvd \echeck{r'}{e} : \estrt{r'} \sreduces i}
{\gvd e : \estrt{r} \sreduces{i} \\ \lang{r'} \subseteq \lang{r}}

\inferline{T-Check-NotOk}
{\gvd \echeck{r'}{e} : \estrt{r'} \sreduces \icheck{r}{i}}
{\gvd e : \estrt{r} \sreduces i \\ \lang{r'} \not \subseteq \lang{r}}

%\inferline{E-dconvert (optional) }
%{\dconvert{\strin[r']}{e} : \strin[r'] \ireduces \ifilter{r'}{\istrf{e}}}
%{e \val }

\end{tabular}
\caption{Type-directed translation to internal terms for $\lcs$}
\label{fig:eeval}
\end{center}
\end{figure}

@@

%\section{Implementation Inside a Python}

@@

>>>>

\documentclass[10pt,preprint]{sigplanconf}
\newcommand{\lamAce}{\lambda_{\text{Ace}}}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{verbatimbox}
\input{../att-icfp14/macros-atlam}
\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{light-gray}{gray}{0.5}
\usepackage{listings}
\usepackage{wasysym}
    \makeatletter
%
% \btIfInRange{number}{range list}{TRUE}{FALSE}
%
% Test if int number <number> is element of a (comma separated) list of ranges
% (such as: {1,3-5,7,10-12,14}) and processes <TRUE> or <FALSE> respectively
%
        \newcount\bt@rangea
        \newcount\bt@rangeb

@@

%
% Begin Grammar 
%
\newcommand{\ereplace}[3]{{\tt replace}[#1](#2,#3)}
\newcommand{\ecoerce}[2]{{\tt coerce}[#1](#2)}
\newcommand{\estr}[1]{{\tt str\_in[#1]}}
\newcommand{\estrt}[1]{{\tt string\_in[#1]}}
\newcommand{\econcat}[2]{{\tt concat}(#1,#2)}

\newcommand{\ireplace}[3]{{\tt ireplace}(#1,#2,#3)}
\newcommand{\icheck}[2]{{\tt icheck}(#1, #2)}
\newcommand{\istr}[1]{{{\tt istr}[#1]}}
\newcommand{\istrt}{{\tt istrt}}
\newcommand{\iconcat}{\hat{+}}
\newcommand{\regext}{ {\tt regex} }

@@

% Remove the < > from grammar productions.
\renewcommand{\grammarlabel}[2]{#1\hfill#2}

@@

<$\psi$> ::=				\hfill	External types					\\
$\estrt{r}$				\hfill Regular expression types \alt

<T> ::= \hfill External expressions \\
      $\estr{s}$ \alt
      $\econcat{e}{e}$ \hfill String concatenation \alt
      $\ereplace{r}{e}{e}$ \alt
      $\ecoerce{r}{e}$

<$\theta$> ::= \hfill Internal types \\
$\istrt$ \alt $\regext$

<S> ::= \hfill Internal expressions \\
  $\istr{s}$ \alt
  $\econcat{i}{i}$ \alt
  $\ireplace{r}{i}{i}$ \alt
  $\icheck{r}{i}$
\end{grammar}
\caption{Syntax for the string sanitation fragment of $\lcs$}
\label{fig:lcsSyntax}
\end{figure}

@@

\onecolumn
\newcommand{\ectx}{?} % Context for external typing
\newcommand{\ictx}{?} % Context for internal typing

\newcommand{\ereduces}{\Downarrow}

\begin{figure}
\begin{center}
\begin{tabular}{c r}

\inferline{Te-include}
{\ectx \vdash \estr{s} : \estrt{r}}
{s \in \lang{r}???}

\inferline{Te-Concat}
{\ectx \vdash \econcat{S_1}{S_2} : \estrt{r_1 \cdot r_2}}
{\ectx \vdash S_1 : \estrt{r_1} \\ \ectx \vdash S_2 : \estrt{r_2}}

\inferline{Te-subst}
{\ectx \vdash \ereplace{r}{S_1}{S_2} : r}
{\ectx \vdash S_1 : \estrt{r_1} \\ \ectx \vdash S_2 : \estrt{r_2} \\ r = {\tt lsubst}(r,r_1,r_2)}

\inferline{Te-coerce}
{\ectx \vdash \ecoerce{r'}{s} : \estrt{r'}}
{\ectx \vdash S : \estrt{r}}

\end{tabular}
\caption{Typing rules for $\lcs$. Err cases?}
\label{fig:etyping}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\begin{tabular}{c r}
\inferline{Te-Include}
{\estr{s} \val}
{ \cdot }

\inferline{Te-Concat}
{\econcat{S_1}{S_2} \ereduces \estr{s_1 \cdot s_2}} % ???
{S_1 \Downarrow \estr{s_1} \\ S_2 \Downarrow \estr{s_2}}

\inferline{Te-Subst}
{\ereplace{r}{S_1}{S_2} \ereduces s}
{S_1 \Downarrow \estr{s_1} \\ S_2 \Downarrow \estr{s_2} \\ s = {\tt substitution}(r,s_1,s_2)}

\inferline{Te-Coerce-OK}
{\ecoerce{r}{s} \ereduces \estr{x}}
{s \ereduces \estr{x} \\ x \in \lang{r}}

\inferline{Te-Coerce-NotOk}
{\ecoerce{r}{s} \ereduces \estr{x} \\ x \not \in \lang{r}}
{s \ereduces \estr{x} \\ x \not \in \lang{r}}
\end{tabular}
\caption{Operational semantics for the external language $\lcs$.
There are also unmentioned error propagation rules for for Te-Concat and Te-Subst.}
\label{fig:eeval}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\begin{tabular}{c r}
\inferline{Ti-Include}
{\istr{s} \val}
{ \cdot }

\inferline{Ti-Include}
{{\tt regex}[r] \val}
{ \cdot }

\inferline{Ti-Concat}
{\iconcat{S_1}{S_2} \ireduces \istr{s_1 \cdot s_2}} % ???
{S_1 \ireduces \istr{s_1} \\ S_2 \ireduces \istr{s_2}}

\inferline{Te-Subst}
{\ireplace{r}{S_2}{S_3} \ireduces {\tt substitution(r,s_2,s_3)}}
{T_1 \ireduces {\tt regex}[r] \\ T_2 \ireduces \istr{s_2} \\ T_3 \ireduces \istr{s_3} \\ }

\inferline{Te-Ccheck-OK}
{\icheck{r}{s} \ireduces \istr{x}}
{s \ireduces \estr{x} \\ x \in \lang{r}}

\inferline{Te-Check-NotOk}
{\icheck{r}{s} \ireduces \istr{x} \\ x \not \in \lang{r}}
{s \ireduces \istr{x} \\ x \not \in \lang{r}}
\end{tabular}
\caption{Operational semantics for the internal language of $\lcs$.
There are also unmentioned error propagation rules for for Te-Concat and Te-Subst.}
\label{fig:ieval}
\end{center}
\end{figure}

%
%\begin{figure}
%\begin{center}
%\begin{tabular}{c r}
%
%\inferline{T-Abs}
%{\gvd \lambda x : \tau .e : \tau \rightarrow \tau' \sreduces \lambda x : \sigma.i}
%{\Gamma, x : \tau \vdash e : \tau' \sreduces i \\ \sigma \equiv \trep(\tau)}
%
%\inferline{T-App}
%{\Gamma \vdash e_1 e_2 \sreduces i_1 i_2}
%{\Gamma \vdash e_1 : \tau \rightarrow \tau' \sreduces i_1 \\ \Gamma \vdash e_2 : \tau \sreduces i_2}
%
%\inferline{T-Replace}
%{\gvd \ereplace{r}{e_1}{e_2} : \estrt{r'} \sreduces \ireplace{r}{i_1}{i_2}}
%{\gvd e_1 : \estrt{r_1} \ireduces i_1 \\ \gvd e_2 : \estrt{r_2} \ireduces i_2 \\ r' \equiv \crl{r}{r_2}{r_1}}
%
%\inferline{T-Check-Ok}
%{\gvd \echeck{r'}{e} : \estrt{r'} \sreduces i}
%{\gvd e : \estrt{r} \sreduces{i} \\ \lang{r'} \subseteq \lang{r}}
%
%\inferline{T-Check-NotOk}
%{\gvd \echeck{r'}{e} : \estrt{r'} \sreduces \icheck{r}{i}}
%{\gvd e : \estrt{r} \sreduces i \\ \lang{r'} \not \subseteq \lang{r}}
%  
%%\inferline{E-dconvert (optional) }
%%{\dconvert{\strin[r']}{e} : \strin[r'] \ireduces \ifilter{r'}{\istrf{e}}}
%%{e \val }
%
%\end{tabular}
%\caption{Type-directed translation to internal terms for $\lcs$}
%\label{fig:eeval}
%\end{center}
%\end{figure}

@@

\section{Implementation Inside a Python}

@@

\newcommand{\F}[1]{\textsf{#1}~}
\newcommand{\FF}[1]{\textsf{#1}}
\newcommand{\Q}{\FF{Arg}}
\newcommand{\xlA}[1]{\lfloor #1 \rfloor_{\lamAce}}
\newcommand{\xA}[1]{$\lfloor #1 \rfloor_{\texttt{ace}}$}
\begin{figure*}
\begin{tabular}{ l r l l }
$\tau$ & & $\xlA{\tau}$ & \xA{\tau}\\
\hline
$\estrt{r}$ & & $\fvar{stringin}[r]$ & \verb|string_in["|$r$\verb|"]|\\
\\
  $e$ & & $\xlA{e}$ & \xA{e}\\
  \hline
  ${\tt str}[r, s]$ & \text{synthetic position} & $\FF{intro}[\FF{str}[s]]()$ : \fvar{stringin}[\FF{rx}[$r$]] & \verb|"|$s$\verb|" (string_in["|$r$\verb|"])|\\
  & \text{analytic position} & $\FF{intro}[\FF{str}[s]]()$ & \verb|"|$s$\verb|"|\\
  ${\tt concat}(e_1; e_2)$ & & $\xlA{e_1}\cdot\FF{elim}[\tvar{concat}](\xlA{e_2})$ & \xA{e_1}\verb| + |\xA{e_2} \\
  ${\tt replace}[r](e_1; e_2)$ & & $\xlA{e_1}\cdot\FF{elim}[\tvar{replace}~\FF{rx}[r]](\xlA{e_2})$ & \xA{e_1}\verb|.replace("|$r$\verb|", |\xA{e_2}\verb|)|\\
  ${\tt coerce}[r](e)$ & & $\xlA{e}\cdot\FF{elim}[\tvar{coerce}~\FF{rx}[r]]()$ & \xA{e}\verb|.coerce("|$r$\verb|")|
\end{tabular}
\caption{Translation of $\lcs$ to $\lamAce$ and Ace.}
\end{figure*}
\begin{figure}
\small\begin{flalign}
& \F{tycon}\fvar{stringin}~\F{of}\FF{R}~\{\\
& \quad \F{iana}\{\tlam{opidx}{\FF{String}}{
	\tlam{tyidx}{\FF{R}}{
	\tlam{a}{\klist{\Q}}{\\
& \quad\quad \tvar{arity0}~\tvar{a}~(\tvar{check}~\tvar{opidx}~\tvar{tyidx}~\titerm{\iup{\tvar{opidx}}})
	}}
}\}\\
& \quad \F{esyn}\{\tlam{opidx}{\kunit+(\FF{R}+\FF{R})}{
	\tlam{a}{\klist{\Q}}{\\
& \quad\quad \tsumcase{\tvar{opidx}}{\_}{\\
& \quad\quad\quad \tvar{arity2}~\tvar{a}~\tlam{a1}{\Q}{\tlam{a2}{\Q}{\\
& \quad\quad\quad\quad \tvar{rsyn}~\tvar{a1}~\tlam{r1}{\FF{R}}{\tlam{i1}{\kITerm}{~}}\\
& \quad\quad\quad\quad \tvar{rsyn}~\tvar{a2}~\tlam{r2}{\FF{R}}{\tlam{i2}{\kITerm}{~}}\\
& \quad\quad\quad\quad\quad (\ttype{stringin}{\FF{rseq}(\tvar{r1}; \tvar{r2})}, \\
& \quad\quad\quad\quad\quad\titerm{{\tt iconcat}(\iup{\tvar{i1}}; \iup{\tvar{i2}})})
}}\\
& \quad\quad}{opidx'}{d}}
}\}
\}
\end{flalign}
\caption{Definition of ...}
\end{figure}
\begin{figure}
\[
\begin{array}{lcl}
% & & Definition & Kind\\
\tvar{concat} & := & \tinl{\kunit, R+R}{b}\\%& \kunit + (R + R)\\
\tvar{replace} & := & \tlam{r}{R}{\tinr{\kunit, R+R}{\tinl{R,R}{\tvar{r}}}}\\% & \karrow{a}{b}\\
\tvar{coerce} & := & \tlam{r}{R}{\tinr{\kunit, R+R}{\tinr{R,R}{\tvar{r}}}}\\
\end{array}
\]
\caption{Definitions}
\end{figure}
